Conflict-Free Vertex Coloring Of Planar Graphs

### 2017-04-03 paper/paper.tex
Six pages has arrived

A graph $G$ is said to be \emph{k-colorable} if it can be colored using $k$ or fewer colors, i.e. $\chi(G) \leq k$. A graph having $\chi(G) = k$ is said to be a \emph{k-chromatic} graph. The \emph{k-colorability} problem (CFCP) asks if a graph can be colored using $k$ colors. This problem is a slightly easier problem than the vertex coloring problem as it is decision problem rather than an optimization problem. An example of a 3-colorable graph and a vertex coloring is shown in figure \ref{fig:vcp-example}.

@@

A \emph{conflict-free k-coloring} of a simple graph $G$ assigns one of $k$ different colors to a subset $S \subseteq V$ of vertices such that $\forall v \in V$, there is a vertex $u \in N(v)$ where the color of $u$ is unique in the closed neighborhood of $v$. Generally, a \emph{conflict-free coloring} of a graph assigns colors to some of the vertices such that, for every vertex $v$, there is a unique color assigned to a vertex among $v$ and $v$'s neighbors. The vertex $u$ can be thought of the \emph{conflict-free neighbor} of $v$. The \emph{conflict-free chromatic number} of G, denoted $\chi_{CF}(G)$, is the smallest $k$ for which a conflict-free coloring exists. \cite{abel2017three}

@@

\subsection{Guaranteeing CF k-Colorability}
It is wasteful to spend time coloring a graph that cannot be conflict-free colored. This leads Abel et al. to provide sufficient criterion to guarantee the conflict-free k-colorability of a certain graphs. To introduce this criterion, we need to introduce a few specific graphs. A complete graph is a simple, undirected graph where every pair of distinct vertices is connected by an edge. A complete graph on $n$ vertices is denoted as $K_n$. The graph $K_n^{-3}$ is the graph obtained by removing any three edges forming a single triangle, i.e. \(K_n^{-3} =  \left ( V(K_n), E(K_n) \setminus E(K_3) \right ) \). A graph $H$ is called a \emph{minor} of a graph $G$ if $H$ can be formed by contracting edges, deleting edges, or deleting vertices. \cite{abel2017three,bondy1976graph}

@@

\subsection{Iterated Elimination of Distance-3-Sets}
The iterated elimination of distance-3-sets algorithm is a polynomial-time heuristic for generating a conflict-free k-coloring of general graphs that satisfy the criterion given in theorem \ref{thm:criterion}. If the criterion is not met, the heuristic will not generate a valid conflict-free k-coloring.

\begin{algorithm}
\caption{Iterated elimination of distance-3-sets} \label{alg:elimination}
\begin{algorithmic}[1]
\State $i \gets 1$
\State Remove all isolated paths from $G$
\While{$G$ is not empty}
	\State $D \gets \emptyset$
	\ForAll{components of $G$}
		\State Pick any vertex $v$
		\State $D \gets D \cup \{ v \}$
		\While{$\exists u$ at distance $\geq 3$ $\forall v \in D$}
			\State Pick $u$ at distance 3 from some vertex in $D$
			\State $D \gets D \cup \{ w \}$
		\EndWhile
		\ForAll{$u \in D$}
			\State Color $u$ with color $i$
		\EndFor
		\State $i \gets i + 1$
		\ForAll{$u \in D$}
			\State Remove $N[u]$ from $G$
		\EndFor
		\State Remove all isolated paths from G
	\EndFor
\EndWhile
\State Color all removed isolated paths using color $i$
\end{algorithmic}
\end{algorithm}

@@

\begin{figure}[h]
	\centering
	\includegraphics[width=6cm]{../figures/algorithm1-step3.pdf}
	\caption{Graph $G_3$ after completing the algorithm}\label{fig:algorithm1-step3}
\end{figure}

@@

It is wise to note that this algorithm does not minimize the number of colored vertices. It can be beneficial to minimize the number of colored vertices (less cost) in certain applications such as building wireless networks. The conflict-free domination number for $k=2$, $\gamma_{CF}^2(G_3)$, is three even though we have four colored vertices in our example.

\section{CF Coloring of Planar Graphs}
\label{sec:planar-coloring}
Conflict-free coloring has some interesting properties when we limit the input graph to be a planar graph. Recall that a graph is planar if it can be drawn on the plane in such a way that its edge intersect only at their endpoints.

\subsection{Bounds for Planar Graphs}
\label{sec:bounds}
Recall the map of the USA example given in the introduction. This is an example of a planar graph and illustrates a famous theorem associated with it. We present this theorem, called the four color theorem, in theorem \ref{thm:four}. It has been proved by multiple groups. The proof by computer that is widely accepted today can be found in \cite{robertson1997four}.

@@

Due to theorem \ref{thm:four}, we immediately know that every planar graph is 4-colorable. This raises the question of whether there are planar graphs that require four colors or if fewer colors could suffice for all planar graphs. This leads us to the conflict-free variation of this theorem. We present this in theorem \ref{thm:four-cf}. It is given and proven in \cite{abel2017three}. In the next subsections, we will investigate this result.

@@

\subsection{One or Two Colors}
\todo[inline]{To do still....}

\subsection{Three or More Colors}
Abel et. al \cite{abel2017three} present specific results for 3-colorable graphs. They examine minimizing the number of colored vertices of 3-colorable graphs. They prove theorems \ref{thm:approximate-cfds} and \ref{thm:outerplanar-cfdn} by using a polynomial-time algorithm that transforms a dominating set $D$ of a planar graph $G$ into a conflict-free k-coloring of $G$ where only the vertices of $D$ are colored. This algorithm is defined in algorithm \ref{alg:cf-dominating}.

\begin{algorithm}
\caption{Conflict-Free Coloring via Dominating Set} \label{alg:cf-dominating}
\begin{algorithmic}[1]
\Input A planar graph $G = (V, E)$
\State Find a dominating set of $G$
\While{$V \setminus D \neq \emptyset$}
	\ForAll{$v \in V \setminus D$}
		\State Pick a vertex $u \in D$
		\State Contract the edge $\{u,v\}$ towards $u$
	\EndFor
\EndWhile
\State Find a proper vertex coloring of $G$
\end{algorithmic}
\end{algorithm}

%To prove theorem \ref{thm:approximate-cfds}, we use a reduction from \emph{k-coloring} to \emph{conflict-free k-coloring}. This reduction is out of the scope of this paper and can be found in \cite{abel2017three}. The reduction preserves the planarity of its input graph, so the resulting graph is still planar. We also know that 3-coloring a graph is NP-complete \cite{sharma2012new}.

Because the planar minimum dominating set problem is NP-hard \cite{garey2002computers}, we know that the conflict-free dominating set problem is NP-hard because we are finding a minimum dominating set. It follows that the corresponding decision problem, the k-conflict-free dominating set problem, is NP-complete.

@@

We get a PTAS for the conflict-free domination number by applying algorithm \ref{alg:cf-dominating} and finding the dominating set using the PTAS defined by Baker and Hill \cite{baker1994approximation}. As the input graph $G$ is planar, the resulting graph $G'$ is also planar because $G'$ is a minor of $G$. Thus, it can be colored with 4 colors by \ref{thm:four}. We then color $G$ with the colors of $G'$, leaving any vertex in $G$ but not in $G'$ uncolored. This leaves us with a conflict-free 4-coloring of $G$. Note that it has a minimal number of vertices colored because we used a PTAS to find the minimum dominating set.

@@

@@

There are many applications of the CFCP that benefit from the recent research into the field. It is easier to minimize frequency assignment for wireless networks and model RFID networks. Current research focuses on finding bounds and properties on specific cases such as planar graphs, interval graphs, types of hypergraphs, and more.

>>>>

A graph $G$ is said to be \emph{k-colorable} if it can be colored using $k$ or fewer colors, i.e. $\chi(G) \leq k$. A graph having $\chi(G) = k$ is said to be a \emph{k-chromatic} graph. The \emph{k-colorability} problem (CFCP) asks if a graph can be colored using $k$ colors. This problem is slightly easier than the VCP as it is decision problem rather than an optimization problem. An example of a 3-colorable graph and a vertex coloring is shown in figure \ref{fig:vcp-example}.

@@

A \emph{conflict-free k-coloring} of a simple graph $G$ assigns one of $k$ different colors to a subset $P \subseteq V$ of vertices such that $\forall v \in V$, there is a vertex $u \in N(v)$ where the color of $u$ is unique in the closed neighborhood of $v$. The set $V \setminus S$ contains any uncolored vertices. Generally, a \emph{conflict-free coloring} of a graph assigns colors to some of the vertices such that, for every vertex $v$, there is a unique color assigned to a vertex among $v$ and $v$'s neighbors. The vertex $u$ can be thought of the \emph{conflict-free neighbor} of $v$. The \emph{conflict-free chromatic number} of G, denoted $\chi_{CF}(G)$, is the smallest $k$ for which a conflict-free coloring exists. \cite{abel2017three}

@@

\subsection{Guaranteeing CF k-Colorability}
It is wasteful to spend time coloring a graph that cannot be conflict-free colored. This leads Abel et al. to provide sufficient criterion to guarantee the conflict-free k-colorability of a certain graphs. To introduce this criterion, we need to introduce a few specific graphs. A complete graph is a simple, undirected graph where every pair of distinct vertices is connected by an edge. A complete graph on $n$ vertices is denoted as $K_n$. The graph $K_n^{-3}$ is the graph obtained by removing any three edges forming a single triangle, i.e. \(K_n^{-3} =  \left ( V(K_n), E(K_n) \setminus E(K_3) \right ) \). A graph $H$ is called a \emph{minor} of a graph $G$ if $H$ can be formed by deleting edges and/or contracting edges (combining the vertices of an edge) from $G$. \cite{abel2017three,bondy1976graph}

@@

\subsection{Iterated Elimination of Distance-3-Sets}
\label{sec:iterated}
The iterated elimination of distance-3-sets algorithm is a polynomial-time heuristic for generating a conflict-free k-coloring of general graphs that satisfy the criterion given in theorem \ref{thm:criterion}. If the criterion is not met, the heuristic will not generate a valid conflict-free k-coloring.

\begin{algorithm}
\caption{Iterated elimination of distance-3-sets} \label{alg:elimination}
\begin{algorithmic}[1]
\Input A simple, undirected graph $G$
\State $i \gets 1$
\State Remove all isolated paths from $G$
\While{$G$ is not empty}
	\State $D \gets \emptyset$
	\ForAll{components of $G$}
		\State Pick any vertex $v$
		\State $D \gets D \cup \{ v \}$
		\While{$\exists u$ at distance $\geq 3$ $\forall v \in D$}
			\State Pick $u$ at distance 3 from some vertex in $D$
			\State $D \gets D \cup \{ w \}$
		\EndWhile
		\ForAll{$u \in D$}
			\State Color $u$ with color $i$
		\EndFor
		\State $i \gets i + 1$
		\ForAll{$u \in D$}
			\State Remove $N[u]$ from $G$
		\EndFor
		\State Remove all isolated paths from G
	\EndFor
\EndWhile
\State Color all removed isolated paths using color $i$
\end{algorithmic}
\end{algorithm}

@@

\begin{figure}[h]
	\centering
	\includegraphics[width=6cm]{../figures/algorithm1-step3.pdf}
	\caption{Coloring of $G_3$ based on algorithm}\label{fig:algorithm1-step3}
\end{figure}

@@

It is wise to note that this algorithm does not necessariyminimize the number of colored vertices. It can be beneficial to minimize the number of colored vertices (less cost) in certain applications such as building wireless networks. The conflict-free domination number for $k=2$ of $G_3$, $\gamma_{CF}^2(G_3)$, is three even though we have four colored vertices in our example. This can be seen by coloring vertices $\{2, 8\}$ with one color and vertex 6 with another.

\section{CF Coloring of Planar Graphs}
\label{sec:planar-coloring}
Conflict-free coloring has some interesting properties when we limit the input graph to be a planar graph. Recall that a graph is planar if it can be drawn on the plane in such a way that its edges intersect only at their endpoints.

\subsection{Bounds for Planar Graphs}
\label{sec:bounds}
Recall the map of the USA example given in the introduction. This is an example of a planar graph and illustrates a famous theorem associated with it. We present this theorem, called the four color theorem, in theorem \ref{thm:four}. It has been proven by multiple groups. The proof by computer that is widely accepted today can be found in \cite{robertson1997four}.

@@

Due to theorem \ref{thm:four}, we immediately know that every planar graph is conflict-free 4-colorable. This raises the question of whether there are planar graphs that require four colors or if fewer colors could suffice for all planar graphs. This leads us to the conflict-free variation of this theorem. We present this in theorem \ref{thm:four-cf}. It is given and proven in \cite{abel2017three}.

@@

\subsection{Minimizing Colored Vertices}
Abel et. al \cite{abel2017three} present specific results for 3-colorable graphs. They examine minimizing the number of colored vertices of 3-colorable graphs. They prove theorems \ref{thm:approximate-cfds} and \ref{thm:outerplanar-cfdn} by using a polynomial-time algorithm that transforms a dominating set $D$ of a planar graph $G$ into a conflict-free k-coloring of $G$ where only the vertices of $D$ are colored. This algorithm is defined in algorithm \ref{alg:cf-dominating}.

\begin{algorithm}
\caption{Conflict-Free Coloring via Dominating Set} \label{alg:cf-dominating}
\begin{algorithmic}[1]
\Input A planar graph $G = (V, E)$
\State Find a dominating set, $D$, of $G$
\ForAll{$v \in V \setminus D$}
	\State Pick a vertex $u \in D$ where $\{u, v\} \in E$
	\State Contract the edge $\{u,v\}$ towards $u$
\EndFor
\State Find a proper vertex coloring of $G$
\State Color the original $G$ with the found coloring
\end{algorithmic}
\end{algorithm}

Because the planar minimum dominating set problem is NP-hard \cite{garey2002computers}, we know that the conflict-free dominating set problem is NP-hard because we aim to find a minimum dominating set. It follows that the corresponding decision problem, the k-conflict-free dominating set problem, is NP-complete.

@@

We get a PTAS for the conflict-free domination number by applying algorithm \ref{alg:cf-dominating} and finding the dominating set using the PTAS defined by Baker and Hill \cite{baker1994approximation}. As the input graph $G$ is planar, the resulting graph $G'$ is also planar because $G'$ is a minor of $G$. Thus, it can be colored with 4 colors by \ref{thm:four}. We then color $G$ with the colors of $G'$, leaving any vertex in $G$ but not in $G'$ uncolored. This leaves us with a conflict-free 4-coloring of $G$.

@@

\subsection{CF Coloring via Dominating Set}
Algorithm \ref{alg:elimination} presented in section \ref{sec:iterated} gives us an efficient heuristic for coloring general graphs including planar graphs. Algorithm \ref{alg:cf-dominating} gives us another way of coloring planar graphs, albeit likely less efficient.

To demonstrate this algorithm, we will utilize graph $G_3$ as we did for algorithm \ref{alg:elimination}. It is a planar graph, so we can use this algorithm to generate a conflict-free coloring. First, we must find a dominating set for $G_3$. Finding a dominating set can be done with a greedy algorithm. For small graphs, such as $G_3$, it is easy to find one by hand. A dominating set of $G_3, \{2, 6, 8\},$ is shown in figure \ref{fig:algorithm2-step1}.

\begin{figure}[h]
	\centering
	\includegraphics[width=6cm]{../figures/algorithm2-step1.pdf}
	\caption{A dominating set of $G_3$}\label{fig:algorithm2-step1}
\end{figure}

We assign our dominating set $\{2,6,8\}$ to $D$. Since $V \setminus D \neq \emptyset$, we enter the \emph{for all} loop in line 2. Here, we iterate through all vertices in $V \setminus D$. For each vertex $v \in V \setminus D$, we will pick another vertex in $D$ that is connected to $v$ by an edge. Then, we will contract the edge $\{u, v\}$ towards $u$. This means that we combine $u$ and $v$ together and update any edges to reflect the changes.

\begin{figure}[h]
	\centering
	\includegraphics[width=6cm]{../figures/algorithm2-step2.pdf}
	\caption{The graph after the first \emph{for} loop}\label{fig:algorithm2-step2}
\end{figure}

To illustrate this, we start the loop by picking a vertex $v$, say vertex 3. We pick vertex $u$ to be vertex 2 because $2 \in D$ and $\{2, 3\} \in E$. We contract the edge and show the resulting graph $G$ in figure \ref{fig:algorithm2-step2}. By iterating over the rest of the vertices in $V \setminus D$, we obtain a graph that is itself a path from vertex 2 to vertex 6 to vertex 8. This graph, along with its proper vertex coloring, is shown in figure \ref{fig:algorithm2-step3}. We can find a proper vertex coloring by any heuristic, such as a greedy algorithm, for the VCP.

\begin{figure}[h]
	\centering
	\includegraphics[width=5cm]{../figures/algorithm2-step3.pdf}
	\caption{The graph after lines 2-5}\label{fig:algorithm2-step3}
\end{figure}

Lastly, we color our original graph $G_3$ based on the proper vertex coloring we generated. We leave any nodes not in the graph $G$ uncolored when we color $G_3$. The coloring of $G_3$ is shown in figure \ref{fig:algorithm2-step4}. It is clear that this has produced a conflict-free coloring. Looking back at algorithm \ref{alg:elimination}, we had 4 colored vertices. Here, we have 3, and it is true that we have minimized colored vertices as $\gamma_{CF}^2(G_3) = 3$.

\begin{figure}[h]
	\centering
	\includegraphics[width=6cm]{../figures/algorithm2-step4.pdf}
	\caption{$G_3$ with minimized colored vertices}\label{fig:algorithm2-step4}
\end{figure}

@@

There are many applications of the CFCP that benefit from the recent research into conflict-free coloring. It is easier to minimize frequency assignment for wireless networks and design RFID networks. Current research focuses on finding bounds and properties on specific cases such as planar graphs, interval graphs, types of hypergraphs, and more.
